Strict Pure Functional Programming and Automatic Parallelization\\ (New-project-in-progess Paper)

### 2016-03-05 pubs/tfp16/paper.tex
sync multithread

% This is LLNCS.DEM the demonstration file of
% the LaTeX macro package from Springer-Verlag
% for Lecture Notes in Computer Science,
% version 2.4 for LaTeX2e as of 16. April 2010
%
\documentclass{llncs}
%
\usepackage{makeidx}  % allows for indexgeneration
%
\begin{document}
%
%mkd \frontmatter          % for the preliminaries
%
\pagestyle{headings}  % switches on printing of running heads
%
%mkd \mainmatter              % start of the contributions
%
\title{Strict Pure Functional Programming and Automatic Parallelization\\
       (New Project-in-progess Paper)}
%
\titlerunning{Automatic Parallelization}  % abbreviated title (for running head)
%                                     also used for the TOC unless
%                                     \toctitle is used
%
\author{Kei Davis\inst{1} \and 
        Dean Prichard\inst{1} \and 
        David Ringo\inst{1,2} \and
        Loren Anderson\inst{1,3} \and
        Jacob Marks\inst{1,4}\thanks{All contributors, past and present, are listed.}
}
%
\authorrunning{Kei Davis et al.} % abbreviated author list (for running head)
%
%%%% list of authors for the TOC (use if author list has to be modified)
%\tocauthor{Ivar Ekeland, Roger Temam, Jeffrey Dean, David Grove,
%Craig Chambers, Kim B. Bruce, and Elisa Bertino}
%
\institute{Los Alamos National Laboratory, Los Alamos, NM, USA\\
\email{kei.davis@lanl.gov},\\ WWW home page:
\texttt{http://ccsweb.lanl.gov/\homedir kei/}
\and
University of New Mexico,
Albuquerque, NM, USA
\and
North Dakota State University,
Fargo, North Dakota, USA
\and 
New Mexico Institute of Mining and Technology,
Socorro, New Mexico, USA
}

@@

@@

In Legion, non-function-local data, or \emph{regions}, are requested from the
runtime system.  Regions may be shared among \emph{tasks} (essentially C++
functions), and each task is prescribed access privileges (read, write,
read/write, etc.)  and coherency requirements for each region to which it will
have access.  Other than registered access to regions tasks must not access
non-constant global data.  In serial execution these access requirements are
notionally superfluous, but in parallel execution the runtime can dynamically
calculate the data dependency graph and relax the programmatically defined
serial ordering of task execution.  Very high parallel efficiency has been
demonstrated on extremely large computing systems~\cite{Bauer14}.  Here at
LANL the Legion system is being strongly considered for a new large-scale C++
code effort.\footnote{It seems significant that the PI of the Legion
  project was previously a co-developer of a strict, arguably pure functional
  language implementation~\cite{AikenFL}.}

@@

\begin{itemize}
\item The implementation must be feasible in terms of effort;
\item The language must be strongly typed, implicitly or explicitly.  More
  specifically, the language should explicitly support both Hindley-Milner
  polymorphism and unboxed types~\cite{firstclass};
\item Straighforward direct linkage with C, preferably without a sophisticated
  foreign function interface, is essential;
\item The language should be \emph{currying-friendly} as described in the ``fast
  curry'' paper~\cite{fastcurry};
\item Choice of degree of strictness is easily selected;
\item Sharing, in the sense of what \emph{lazy} means vs.\ merely non-strict,
  should be preserved for all degrees of strictness;
\item Tail calling should be properly implemented, i.e., not as a trampoline;
\item etc.
\end{itemize}

@@

@@

Other possible variations concern the notional arity of the underlying
function being applied.  This is readily explained in terms of the operation
of the STG machine described later.  Our description is based on that given in
Marlow and Peyton Jones' \emph{fast curry} paper~\cite{fastcurry}, to which we
refer so many times that we subsequenty refer to it as
\emph{EA}.\footnote{\emph{Eval/Apply} from the title because \emph{FC} likely has another
  connotation to the reader.}

In STG the arity of a function is syntactically defined as the number of
manifest formal parameters, thus \texttt{id x = x} has by definition arity 1
regardless of what type it might be dynamically ``instantiated'' at.  We distinguish
\emph{application}, e.g.\ \texttt{f x y}, which is a syntactic notion, from
actually calling a function. Thus \texttt{id id x} is an application of \texttt{id}
to two arguments, but the underlying \emph{FUN} (user- or system- defined function)
when actually called is given exactly one argument.

@@

Proper tail calling is achieved.  Because we maintain our own data and control
(continuation) stack all generated C functions take no arguments and have
return type void, and with a single exception in the non-strict case are
invoked in tail call positions, both gcc and Clang/LLVM generate jumps instead
of calls for C calls in tail call positions.

@@

At the time of writing we have a complete serial implementation: STG parser,
type inferencer, code generator, and runtime system including a garbage
collector.

@@

DOE/NNSA/LANL LDRD, DOE/NNSA SULI.

%
% ---- Bibliography ----
%
\begin{thebibliography}{5}

\bibitem{Heroux}
January 14, 2016
Exascale Programming: Adapting What We Have Can (and Must) Work
Michael A. Heroux, Sandia National Laboratories
http://www.hpcwire.com/2016/01/14/24151/

\bibitem{EUexascale}
Toward Exascale Computing.
European Exascale Software Initiative.
http://www.eesi-project.eu/

\bibitem{Davis96}
MPP Parallel Haskell. 
K. Davis. 
Draft Proceedings Implementation of Functional Languages 1996 (IFL '96) LNCS
1268, pp49-54, Bonn/Bad-Godesberg, Germany, LNCS 1268, Springer-Verlag.

\bibitem{Bauer12} 
Legion: Expressing Locality and Independence with Logical
  Regions. M. Bauer, S. Treichler, E. Slaughter and A. Aiken. In Proceedings
  of the Conference on Supercomputing, pages 1-11, November 2012.

\bibitem{Bauer14} 
Structure Slicing: Extending Logical Regions with Fields. M. Bauer, S. Treichler, E. Slaughter and A. Aiken. In Proceedings of the Conference on Supercomputing, pages 845-856, November 2014.

\bibitem{AikenFL}
A. Aiken, J.H. Williams, and E.L. Wimmers
The FL Project:  The Design of a Functional Language
(technical report not published elsewhere), September 1993.
http://theory.stanford.edu/~aiken/publications/trs/FLProject.pdf

\bibitem{Bird84}
R. S. Bird, 
Using circular programs to eliminate multiple traversals of data, 
Acta Informatica, 21(3), 239-250 (1984).

\bibitem{Johnsson87}
Thomas Johnsson
Attribute Grammars as a Functional Programming Paradigm
Proc. of a conference on Functional programming languages and computer architecture
Pages 154-173 
Springer-Verlag London, UK ©1987 
table of contents ISBN:0-387-18317-5

\bibitem{Hughes89}
J. Hughes,
Why Functional Programming Matters,
Computer Journal,
32,
2,
98--107,
1989.

\bibitem{strict-strictdata}
http://ghc.haskell.org/trac/ghc/wiki/StrictPragma

[ghclib] GHC/As a library http://wiki.haskell.org/GHC/As\_a\_library

\end{thebibliography}

>>>>

% This is LLNCS.DEM the demonstration file of
% the LaTeX macro package from Springer-Verlag
% for Lecture Notes in Computer Science,
% version 2.4 for LaTeX2e as of 16. April 2010
%
\documentclass{llncs}
%
\usepackage{makeidx}  % allows for indexgeneration
%
\begin{document}
%
%mkd \frontmatter          % for the preliminaries
%
\pagestyle{headings}  % switches on printing of running heads
%
%mkd \mainmatter              % start of the contributions
%
\title{Strict Pure Functional Programming and Automatic Parallelization\\
       (New-project-in-progess Paper)}
%
\titlerunning{Automatic Parallelization}  % abbreviated title (for running head)
%                                     also used for the TOC unless
%                                     \toctitle is used
%
\author{Kei Davis\inst{1} \and 
        Dean Prichard\inst{1} \and 
        David Ringo\inst{1,2} \and
        Loren Anderson\inst{1,3} \and
        Jacob Marks\inst{1,4}\thanks{All contributors, past and present, are listed.}
}
%
\authorrunning{Kei Davis et al.} % abbreviated author list (for running head)
%
%%%% list of authors for the TOC (use if author list has to be modified)
%\tocauthor{Ivar Ekeland, Roger Temam, Jeffrey Dean, David Grove,
%Craig Chambers, Kim B. Bruce, and Elisa Bertino}
%
\institute{Los Alamos National Laboratory, Los Alamos, NM, USA\\
\email{kei.davis@lanl.gov},\\ WWW home page:
\texttt{http://ccsweb.lanl.gov/\homedir kei/}
\and
University of New Mexico,
Albuquerque, NM, USA
\and
North Dakota State University,
Fargo, North Dakota, USA
\and 
New Mexico Institute of Mining and Technology,
Socorro, New Mexico, USA
}

@@

\emph{Note 3: Hopefully in the spirit of a student-friendly workshop, in this
  draft we have chosen a more conversational style of presentation than
  necessarily appropriate for a final paper, and have also highlighted the
  major student contributions, and perhaps gratuitous detail about process.}

@@

In Legion, non-function-local data, or \emph{regions}, are requested from the
runtime system.  Regions may be shared among \emph{tasks} (essentially C++
functions), and each task is prescribed access privileges (read, write,
read/write, etc.)  and coherency requirements for each region to which it will
have access.  Other than registered access to regions tasks must not access
non-constant global data.  In serial execution these access requirements are
notionally superfluous, but in parallel execution the runtime can dynamically
calculate the data dependency graph and relax the programmatically defined
serial ordering of task execution.  Very high parallel efficiency has been
demonstrated on extremely large computing systems~\cite{Bauer14}.  Here at
LANL the Legion system is being strongly considered for a new large-scale C++
code effort.\footnote{It seems significant that the PI of the Legion
  project was previously a co-developer of a strict, arguably pure functional
  language implementation~\cite{AikenFL,FLreport89}, in conjunction with
John Backus as a refinement of his FP language~\cite{Backus:1978}.}

@@

\begin{itemize}
\item The implementation must be feasible in terms of effort;
\item The language must be strongly typed, implicitly or explicitly.  More
  specifically, the language should explicitly support both Hindley-Milner
  polymorphism and unboxed types~\cite{Jones:1991};
\item Straighforward direct linkage with C, preferably without a sophisticated
  foreign function interface, is essential;
\item The language should be \emph{currying-friendly} as described by Marlow
and Peyton Jones~\cite{Marlow2004};
\item Choice of degree of strictness is easily selected;
\item Sharing, in the sense of what \emph{lazy} means vs.\ merely non-strict,
  should be preserved for all degrees of strictness;
\item Tail calling should be properly implemented, i.e., not as a trampoline;
\item etc.
\end{itemize}

@@

Operationally, when an application such as

@@

Other possible variations concern the notional arity of the underlying
function being applied.  This is readily explained in terms of the operation
of the STG machine described later.  Our description is based on that given in
Marlow and Peyton Jones' \emph{fast curry} paper~\cite{fastcurry}, to which we
refer so many times that we will subsequenty refer to it as
\emph{EA}.\footnote{\emph{Eval/Apply} from the title because \emph{FC} likely
  has another connotation to the reader.}

In STG the arity of a function is syntactically defined as the number of
manifest formal parameters, thus \texttt{id x = x} has by definition arity 1
regardless of what argument type it might be dynamically applied to.  We
distinguish \emph{application}, e.g.\ \texttt{f x y}, which is a syntactic
notion, from actually calling a function. Thus \texttt{id id x} is an
application of \texttt{id} to two arguments, but the underlying \emph{FUN}
(user- or system- defined function) when actually called is given exactly one
argument.

@@

Proper tail calling is achieved.  Because we maintain our own data and control
(continuation) stack all generated C functions take no arguments and have
return type void, and with a single exception in the non-strict case are
invoked in tail call positions, at non-zerop optimization levels both gcc and
Clang/LLVM generate jumps instead of calls for C calls in tail call positions.

@@

At the time of writing we have a complete serial implementation, including
\begin{itemize}
\item STG (inc. ADT w/unboxed) parser
\item type inferencer
\item code generator
\item runtime system including a garbage collector
\end{itemize}

@@

Funding for this project and its participants has been provided by the DOE
NNSA LANL Laboratory Directed Research and Development program award
20150845ER, the National Science Foundation Science, Technology, Engineering,
and Mathematics Talent Expansion Program (STEP) program for undergraduate
students, and the DOE Science Undergraduate Laboratory Internship program for
undergraduate students.
Los Alamos National Laboratory is managed and operated by Los Alamos National
Security, LLC (LANS), under contract number DE-AC52-06NA25396 for the
Department of Energy’s National Nuclear Security Administration (NNSA).

% just get these into the bib
\cite{ghc-as-library}
\cite{Heren02}

%
% ---- Bibliography ----
%

\bibliographystyle{splncs03}

\bibliography{tfp}
